Estimation of the Effect of a Continuous Treatment Variable

### 2025-11-19 presentation/proposal/proposal.qmd
presentation for proposal defense added

---
title: "Chapter 1: A new model selection approach based on local economically optimal input rate"
author:
  - name: Taro Mieno$^*$
  - name: Mona Mousavi$^*$
  - name: David Bullock$^\dagger$
institute:
  - $^*$ University of Nebraska Lincoln
  - $^\dagger$ University of Illinois Urbana Champaign 
date: 06/12/2025
format:
  beamer:
    theme: Singapore
    navigation: horizontal
    section-titles: false
    slide-level: 2
    fontsize: "10pt"
header-includes: 
- \logo{\ifnum\thepage>1\includegraphics[width=1cm]{nebraska-n.jpg}\fi}
---

@@

@@

+ Variable-rate input management as a solution (accounts for spatial heterogeneity in crop input demand.

@@

## The simulated OFPE field:
```{r}
#| echo: false
#| fig-align: "center"
#| out-width: "95%"
knitr::include_graphics("~/Desktop/g_field_map.png")
```

@@

```{r}
#| echo: false
#| fig-align: "center"
#| out-width: "100%"
knitr::include_graphics("g_N.png")
```

@@

```{r}
#| echo: false
#| fig-align: "center"
#| out-width: "100%"
knitr::include_graphics("/Users/monamousavi/Dropbox/Causal_climate/g_y.png")
```

@@

+ Steps for each fold and candidate models:
  1. train the GAM model using the **test** data 
  2. estimate uniform EONR for the **test** data ($\hat{N}^*_{f,gam}$)
  3. train the candidate model using the **train** data 
  4. estimate site-specific EONR for the **test** data based on the trained candidate model ($\hat{N}^*_{i, f,m}$). 
  5. Calcurate the average of the site-specific EONR ($\hat{N}^*_{f,m}$)
  6. Calcurate the RMSE of the estimated local EONR:

@@

>>>>

---
title: "Proposal Defense"
author:
  - name: Mona Mousavi$^*$
institute:
  - $^*$ University of Nebraska Lincoln
date: 06/12/2025
format:
  revealjs: default
  beamer:
    theme: Singapore
    navigation: horizontal
    section-titles: false
    slide-level: 2
    fontsize: "10pt"
header-includes: 
logo: nebraska-n.jpg
---

@@

# Chapter 1: A new model selection approach based on local economically optimal input rate

@@

+ Variable-rate input management as a solution (accounts for spatial heterogeneity in crop input demand).

@@

## The simulated OFPE field:
```{r}
#| echo: false
#| fig-align: "center"
#| out-width: "95%"
knitr::include_graphics(
  here::here("presentation/proposal/g_field_map.png")
)
```

@@

```{r}
#| echo: false
#| fig-align: "center"
#| out-width: "100%"
knitr::include_graphics(
  here::here("presentation/proposal/g_N.png")
)
```

@@

```{r}
#| echo: false
#| fig-align: "center"
#| out-width: "100%"
knitr::include_graphics(
  here::here("presentation/proposal/g_y.png")
)
```

@@

+ Steps for each fold and candidate models:
  1. train the GAM model using the **test** data 
  2. estimate uniform EONR for the **test** data ($\hat{N}^*_{f,gam}$)
  3. train the candidate model using the **train** data 
  4. estimate site-specific EONR for the **test** data based on the trained candidate model ($\hat{N}^*_{i, f,m}$). 
  5. Calculate the average of the site-specific EONR ($\hat{N}^*_{f,m}$)
  6. Calculate the RMSE of the estimated local EONR:

@@

# Chapter 3: Economic assessment of biochar as a means to enhance soil water holding capacity

+ Improving soil health is considered to be an important means of enhancing crop production resilience to climate change
+ Soil water holding capacity (WHC) is an important property of soil. Higher soil WHC leads to
  + withstanding droughts better
  + less irrigation (conserving water resources)
+ Biochar is actively researched as a supplement to enhance soil WHC of production fields, but is expensive

## Research question and objective

:::{.callout-note title="Research Question"}
+ Is biochar economically viable?
+ If not, how much cost reduction in biochar needs to be achieved to make it economically viable?
:::

1. Simulate corn production using APSIM (a crop growth model) at different values of soil WHC for
   1. historical weather conditions 
   2. weather scenarios under climate change
2. Calculate profit
3. Identify the value of soil WHC enhancement

## Expected Outcomes

### 2025-11-19 writing/manuscript_CF_continuous.qmd
presentation for proposal defense added

The key objective of this article is to identify a causal machine learning approach capable of estimating heterogeneous treatment effects for yield response function estimation using data from on-farm trials, where the treatment variable—nitrogen rate—is continuous and randomized. We investigate two complementary approaches: a Single-Orthogonalized Neural Network (SO-ANN), designed to flexibly capture nonlinear treatment–response relationships, and an extended Causal Forest (CF) that incorporates smoothing splines to model continuous treatments. Both methods aim to preserve the strengths of causal ML—nonparametric modeling of complex interactions without restrictive assumptions. Their performance is evaluated through Monte Carlo simulations that mimic typical OFPE conditions and crop yield responses to nitrogen.

::: {.column-margin}
<span style = "color: blue;"> We need to emphasize that we are proposing approaches that have not been tried anywhere: SO-NN and semi-parametric-CF. They are the main approaches. We then have other contrasting approaches to compare against. </span>
:::
<!--end of column-margin-->

@@

::: {.column-margin}
<span style = "color: blue;"> SO-ANN not introduced yet. You need to talk more about what you do in the above paragraph (mostly SO-NN and semi-parametric CF). </span>
:::
<!--end of column-margin-->

@@

::: {.column-margin}
<span style = "color: blue;"> Explain in  more detail about why it this is innovative. Explain the concept of linearization with splines using the figure I generated in my old manuscript. </span>
:::
<!--end of column-margin-->

@@

$$
\begin{align} 
y_{i,j} = 
  \begin{cases}
  \alpha_{i,j} + \beta_{i,j} N + \gamma_{i,j} N^2 + \varepsilon_{i,j}, & N < \tau_{i,j} \\
  \eta_{i,j} + \varepsilon_{i,j}, & N \geq \tau_{i,j}
  \end{cases}
\label{eq:yield-gen-sim}
\end{align}
$$

>>>>

The objective of this study is to develop and evaluate new causal machine-learning approaches capable of estimating heterogeneous treatment effects for yield response function estimation using data from on-farm trials, where the treatment variable—nitrogen rate—is continuous and randomized. We introduce two methods: a Single-Orthogonalized Neural Network (SO-NN), which isolates the treatment–response component by removing variation in yield explained solely by field conditions, and a semi-parametric Causal Forest (CF) that incorporates smoothing splines to accommodate continuous treatments. Both methods preserve key strengths of causal machine learning by flexibly modeling complex interactions without imposing restrictive functional assumptions. To assess their performance, we compare them with several existing approaches, such as direct applications of random forests or neural networks, using a Monte Carlo simulation framework.

@@

@@

@@

$$
\begin{aligned}
y_{i,j} =
  \begin{cases}
    \alpha_{i,j} + \beta_{i,j} N + \gamma_{i,j} N^2 + \varepsilon_{i,j}, & N < \tau_{i,j} \\
    \eta_{i,j} + \varepsilon_{i,j}, & N \ge \tau_{i,j}
  \end{cases}
\end{aligned}
\label{eq:yield-gen-sim}
$$
